\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}

### 2021-10-15 Manuscripts/Manuscript by chapters- SIM Revisions/00-Full_document_No_Appendix.tex
Working on reviewers' comments

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Using generalized additive models to analyze biomedical non-linear longitudinal data},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\newcommand{\passthrough}[1]{#1}
\lstset{defaultdialect=[5.3]Lua}
\lstset{defaultdialect=[x86masm]Assembler}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{helvet}
\usepackage{lineno}
\usepackage[affil-it,blocks]{authblk}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage[nomarkers,figuresonly]{endfloat}
%set a box to put the ORCID logo
\newbox{\myorcidaffilbox}
\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{orcid}}

@@

%line spacing
\doublespacing
\usepackage{placeins}
\usepackage{subfig}
\usepackage{breqn}
\usepackage[font={small}]{caption}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
\date{\vspace{-2.5em}}

@@

Splines (cubic, thin plate, etc.) are commonly used \emph{basis functions}; a cubic spline is a smooth curve constructed from cubic polynomials joined together in a manner that enforces smoothness, and thin plate regression splines are an optimized version that work well with noisy data.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37}} Splines have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{53}} Therefore, this data-driven flexibility in GAMs overcomes the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear.

To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C. The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time. The timeline can be divided in equally spaced \emph{knots}, each knot being a region where a different set of basis functions will be used. Because there are six timepoints for this group, five knots can be used. The model with five knots to construct the smooth term means that it will have four basis functions (plus one that corresponds to the intercept). The choice of basis functions is set using default values in the package \emph{mgcv} depending on the number of knots. In Figure \ref{fig:basis-plot}A, the four basis functions (and the intercept) are shown. Each of the basis functions is composed of six different points (because there are six points on the timeline). To control the ``wiggliness'' of the fit, each of the basis functions of Figure \ref{fig:basis-plot}A is weighted by multiplying it by a coefficient according to the matrix of Figure \ref{fig:basis-plot}B. The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the ``wiggliness'' of the smooth fit to prevent overfitting. A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.

To get the weighted basis functions, each basis (from Figure Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions. Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions. Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is less than 1). On the other hand, the third basis function has roughly doubled its value. Finally, the weighted basis functions are added at each timepoint to produce the smooth term. The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (orange line), along the simulated values per group, which appear as points.

@@

\caption{Basis functions for a single smoother for time with five knots. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2. B: Matrix of basis function weights. Each basis function is multiplied by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Weighted basis functions. Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B. Note the corresponding increase (or decrease) in magnitude of each weighted basis function. D: Smoother for time and original data points. The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot}
\end{figure}

@@

Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{54}} In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case).\textsuperscript{\protect\hyperlink{ref-miller2019}{55}} Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model.\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{56}} Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model. In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{55}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{57}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.

@@

For the data in Figure \ref{fig:sim-smooth-plot}, A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment. To estimate this, the model incorporates independent smooths for \emph{Group} and \emph{Day}, respectively. The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (\emph{Day}) and the other is a factor (\emph{Group}). Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the groups. Using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!} the model syntax is:

\begin{lstlisting}[language=R]
m1 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
\end{lstlisting}

This syntax specifies that \passthrough{\lstinline!m1!} will store the model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots. The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B). Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{58}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{59}}

@@

Biomedical longitudinal non-linear data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA. Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \ref{simulation}. This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model. Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{60}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.

In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data. It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{61}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.

@@

We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible. In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{62},\protect\hyperlink{ref-weissgerber2018}{63}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field. Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research. We have made all the data and code used in this paper accessible, and we hope that this will encourage other researchers to do the same with future projects.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\hypertarget{ref-roblyer2011}{}%
\CSLLeftMargin{1. }
\CSLRightInline{Roblyer D, Ueda S, Cerussi A, et al. Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment. \emph{Proceedings of the National Academy of Sciences}. 2011;108(35):14626-14631. doi:\href{https://doi.org/10.1073/pnas.1013103108}{10.1073/pnas.1013103108}}

\leavevmode\hypertarget{ref-tank2020}{}%
\CSLLeftMargin{2. }
\CSLRightInline{Tank A, Peterson HM, Pera V, et al. Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens. \emph{Breast Cancer Research}. 2020;22(1). doi:\href{https://doi.org/10.1186/s13058-020-01262-1}{10.1186/s13058-020-01262-1}}

\leavevmode\hypertarget{ref-pavlov2018}{}%
\CSLLeftMargin{3. }
\CSLRightInline{Pavlov MV, Kalganova TI, Lyubimtseva YS. Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy. \emph{Journal of Biomedical Optics}. 2018;23(09):1. doi:\href{https://doi.org/10.1117/1.jbo.23.9.091410}{10.1117/1.jbo.23.9.091410}}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
\date{\vspace{-2.5em}}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Using generalized additive models to analyze biomedical non-linear longitudinal data},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\newcommand{\passthrough}[1]{#1}
\lstset{defaultdialect=[5.3]Lua}
\lstset{defaultdialect=[x86masm]Assembler}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{mathptmx}
\usepackage{lineno}
\usepackage[affil-it,blocks]{authblk}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{setspace}
%\usepackage[nomarkers,figuresonly]{endfloat}
%set a box to put the ORCID logo
\newbox{\myorcidaffilbox}
\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{orcid}}

@@

%line spacing
\doublespacing
\usepackage{placeins}
\usepackage{subfig}
\usepackage{breqn}
\usepackage[font={small}]{caption}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{53}} Although different types of splines exist, cubic and thin plate regression splines will be briefly discussed next to give a general idea of the development within the GAM framework.

Cubic splines (CS), are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness. The use of CS as smoothers in GAMs was discussed within the original GAM framework,\textsuperscript{\protect\hyperlink{ref-hastie1987}{51}} but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit.\textsuperscript{\protect\hyperlink{ref-wood2003}{54}} In contrast, Wood\textsuperscript{\protect\hyperlink{ref-wood2003}{54}} provided a computationally efficient implementation of thin plate regression splines(TPRS), which do not require the knot placement, and require simply a choice on the number of basis to be used to create the smoother (for mathematical details see refs. \normalfont{[@wood2003]} and).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}

\textbf{To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \ref{fig:l-q-response}C. The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time. A computational requisite is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable. Because the data has 6 unique time points, we can specify a maximum of six basis functions to create the smooth (it is important to note that is not necessary to specify an equal number of basis to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data).}

If five basis functions are used to fit a GAM for the data that appears in Figure \ref{fig:l-q-response}C, the resulting fitting process is shown in Figure \ref{fig:basis-plot}A. The four basis functions (and the intercept) are shown. Each of the basis functions is composed of six different points (because there are six points on the timeline). To control the ``wiggliness'' of the fit, each of the basis functions of Figure \ref{fig:basis-plot}A is weighted by multiplying it by a coefficient according to the matrix of Figure \ref{fig:basis-plot}B. The parameter estimates are penalized (shrunk towards 0) where the penalty reduces the ``wiggliness'' of the smooth fit to prevent overfitting. A weak penalty estimate will result in wiggly functions whereas a strong penalty estimate provides evidence that a linear response is appropriate.

To get the weighted basis functions, each basis (from Figure Figure \ref{fig:basis-plot}A) is multiplied by the corresponding coefficients in Figure \ref{fig:basis-plot}B, thereby increasing or decreasing the original basis functions. Figure \ref{fig:basis-plot}C shows the resulting weighted basis functions. Note that the magnitude of the weighting for the first basis function has resulted in a decrease of its overall value (because the coefficient for that basis function is \textbf{negative} and less than 1). On the other hand, the third basis function has roughly doubled its value. Finally, the weighted basis functions are added at each timepoint to produce the smooth term. The resulting smooth term for the effect of \emph{time} is shown in Figure \ref{fig:basis-plot}D (orange line), along the simulated values per group, which appear as points.

@@

\caption{Basis functions for a single smoother for time. A: Basis functions for a single smoother for time for the simulated data of Group 1 from Figure 2. B: Matrix of basis function weights. Each basis function is multiplied by a coefficient which can be positive or negative. The coefficient determines the overall effect of each basis in the final smoother. C: Weighted basis functions. Each of the four basis functions of panel A has been weighted by the corresponding coefficient shown in Panel B. Note the corresponding increase (or decrease) in magnitude of each weighted basis function. D: Smoother for time and original data points. The smoother (line) is the result of the sum of each weighted basis function at each time point, with simulated values for the group shown as points.}\label{fig:basis-plot}
\end{figure}

@@

Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{55}} In the case of non-linear data, the belief that the \emph{true} trend of the data is likely to be smooth rather than ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} GAMs are considered ``empirical'' Bayesian models because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``Full Bayes'' case).\textsuperscript{\protect\hyperlink{ref-miller2019}{56}} Moreover, the use of the restricted maximum likelihood (REML) to estimate the smoothing parameters gives an empirical estimate of the smooth model.\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-laird1982}{57}} Therefore, the confidence intervals calculated for the smooth terms using the package \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (pointwise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} This last part means that contrary to a pointwise coverage (where the coverage of the interval is correct for a single point) the estimated confidence intervals for the smooths will contain \emph{on average} the true function of the data 95\% of the time across the entire timeline (in the case of longitudinal data for which smooths are calculated), which allows to obtain better inference from the model. In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{56}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{58}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.

@@

For the data in Figure \ref{fig:sim-smooth-plot}, A the main effect of interest is how \(\mbox{StO}_2\) changes over time for each treatment. To estimate this, the model incorporates independent smooths for \emph{Group} and \emph{Day}, respectively. The main thing to consider is that model syntax accounts for the fact that one of the variables is numeric (\emph{Day}) and the other is a factor (\emph{Group}). Because the smooths are centered at 0, the factor variable needs to be specified as a parametric term in order to identify any differences between the group means. Using \(\textsf{R}\) and the package \passthrough{\lstinline!mgcv!} the model syntax is:

\begin{lstlisting}[language=R]
gam_02 <- gam(StO2_sim ~ Group + s(Day, by=Group, k=5), method='REML', data = dat_sim)
\end{lstlisting}

This syntax specifies that \passthrough{\lstinline!gam\_02!} (named accordingly so it matches the model sequence from the Appendix) contains the fitted model, and that the change in the simulated oxygen saturation (\passthrough{\lstinline!StO2\_sim!}) is modeled using independent smooths over \passthrough{\lstinline!Day!} for each \passthrough{\lstinline!Group!} (the parenthesis preceded by \passthrough{\lstinline!s!}) using 5 knots. The smooth is constructed by default using thin plate regression splines, but other splines can be used if desired, including Gaussian process smooths.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} The parametric term \passthrough{\lstinline!Group!} is added to quantify overall mean differences in the effect of treatment between groups, and the \passthrough{\lstinline!method!} chosen to estimate the smoothing parameters is the restricted maximum likelihood (REML).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B). Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{59}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{60}}

@@

Biomedical longitudinal non-linear data is particularly challenging to analyze due to the likelihood of missing observations and different correlation structures in the data, which limit the use of rm-ANOVA. Although LMEMs have started to replace rm-ANOVA as the choice to analyze biomedical data, both methods yield biased estimates when they are used to fit non-linear data as we have visually demonstrated in Section \ref{simulation}. This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model. Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{61}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, instead of merely relying on a \(R^2\) value.

In this paper we have presented GAMs as a suitable method to analyze non-linear longitudinal data. It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{52}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{62}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the theory quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of following non-linear trends in the data.

@@

We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible. In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{63},\protect\hyperlink{ref-weissgerber2018}{64}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field. Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research. We have made all the data and code used in this paper accessible, and we hope that this will encourage other researchers to do the same with future projects.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-roblyer2011}{}}%
\CSLLeftMargin{1. }
\CSLRightInline{Roblyer D, Ueda S, Cerussi A, et al. Optical imaging of breast cancer oxyhemoglobin flare correlates with neoadjuvant chemotherapy response one day after starting treatment. \emph{Proceedings of the National Academy of Sciences}. 2011;108(35):14626-14631. doi:\href{https://doi.org/10.1073/pnas.1013103108}{10.1073/pnas.1013103108}}

\leavevmode\vadjust pre{\hypertarget{ref-tank2020}{}}%
\CSLLeftMargin{2. }
\CSLRightInline{Tank A, Peterson HM, Pera V, et al. Diffuse optical spectroscopic imaging reveals distinct early breast tumor hemodynamic responses to metronomic and maximum tolerated dose regimens. \emph{Breast Cancer Research}. 2020;22(1). doi:\href{https://doi.org/10.1186/s13058-020-01262-1}{10.1186/s13058-020-01262-1}}

\leavevmode\vadjust pre{\hypertarget{ref-pavlov2018}{}}%
\CSLLeftMargin{3. }
\CSLRightInline{Pavlov MV, Kalganova TI, Lyubimtseva YS. Multimodal approach in assessment of the response of breast cancer to neoadjuvant chemotherapy. \emph{Journal of Biomedical Optics}. 2018;23(09):1. doi:\href{https://doi.org/10.1117/1.jbo.23.9.091410}{10.1117/1.jbo.23.9.091410}}

\leavevmode\vadjust pre{\hypertarget{ref-demidov2018}{}}%
\CSLLeftMargin{4. }
\CSLRightInline{Demidov V, Maeda A, Sugita M, et al. Preclinical longitudinal imaging of tumor microvascular radiobiological response with functional optical coherence tomography. \emph{Scientific Reports}. 2018;8(1). doi:\href{https://doi.org/10.1038/s41598-017-18635-w}{10.1038/s41598-017-18635-w}}
